Testing and Isolation Efficacy: Insights from a Simple Epidemic Model

### 2021-06-03 SIR_manuscript.tex
slightly improve make rule for modeldefs
the rho range in high testing was above omega, caused numerical artifact, fixed now
meeeting with ali

%% FIXME: uncomment when this pipeline is working
%% \input{codes/modeldefs.tex}

@@

%% FIXME: uncomment when pipeline works
%% $\beta=\betaparam$; $\gamma=\gammaparam$

@@

\newpage
\begin{figure}[h!]
\centering
\begin{subfigure}[t]{.46\textwidth}
\centering
\includegraphics[width=\linewidth]{codes/R0contour_random.pdf}
\caption{}\label{p.a}
\end{subfigure}
%
\begin{subfigure}[t]{.44\textwidth}
\centering
\includegraphics[width=\linewidth]{codes/R0contour_TTI.pdf}
\caption{}\label{p.b}
\end{subfigure}
\caption{
{\bf A comparison of effectiveness control on $\Rnum$ at low level of \percap testing intensity $\rho$.}
Numerical evaluation of the effectiveness control parameter, $\Delta$ (Eq.~\ref{eq:del4}), with respect to changes in testing and isolation. Parameter values (Table~\ref{tab:params}):
$\beta=0.25$ day$^{-1}$, $1/\gamma= 1/3$ days; $\omega \in [1/12,2]$ day$^{-1}$;  $\rho \in [0,0.013]$ day$^{-1}$; $\theta\_w$ and $\theta\_c$ vary between 0 (no effect of isolation) and 1 (complete elimination of transmission); $p_S=0$, $p_I=1$ and $p_R=0.5$. Contours of $\Delta$ are plotted for (a) random testing ($w_S=w_I=w_R=1$) and (b) targeted testing ($w_S=0.3$; $w_I=w_R=1$). 
The baseline basic reproduction number $\Rnum$ and the doubling time $r$, i.e. $\beta/\gamma$ and $\beta-\gamma$ respectively, are approximately 1.5 and 0.08 in this simulation.
}
\label{pan}
\end{figure}

\begin{figure}[h!]
\centering
\begin{subfigure}[t]{.46\textwidth}
\centering
\includegraphics[width=\linewidth]{codes/R0contour_random2.pdf}
\caption{}
\end{subfigure}
%
\begin{subfigure}[t]{.44\textwidth}
\centering
\includegraphics[width=\linewidth]{codes/R0contour_TTI2.pdf}
\caption{}
\end{subfigure}
\caption{
{\bf A comparison of effectiveness control on $\Rnum$ at high level of \percap testing intensity $\rho$.}
We numerically evaluate the effectiveness control parameter, $\Delta$ (Eq.~\ref{eq:del4}), with respect to testing and isolation. We use the following parameters (see Table~\ref{tab:params} for a brief description):
$\omega \in [1/5,2]$ 1/day, $1/\gamma= 6$ days, $\rho \in [0,1/5)$ 1/day, $\theta\_w$ and $\theta\_c$ vary between 0 and 1 with 0 for no effect and 1 for full effect of isolation on the transmission probability, $\beta=0.25$ 1/day, $p_S=0$, $p_I=1$ and $p_R=0.5$. Contours of $\Delta$ are plotted for two testing strategies identified by a set of relative testing weights; (a) random testing where $w_S=w_I=w_R=1$ and (b) targeted testing where $w_S=0.3$ and $w_I=w_R=1$. The baseline basic reproduction number $\Rnum$ and the doubling time $r$, i.e. $\beta/\gamma$ and $\beta-\gamma$ respectively, are approximately 1.5 and 0.08 in this simulation.
}
\label{pan2}
\end{figure}

@@

\begin{itemize}
\item When waiting individuals do not isolate (e.g. $\theta_w =0$, top row of \fref{pan}), increasing testing delays (i.e. decreasing $\omega$) increases \Rnum, as we might expect: $\Delta$ decreases --- a larger region of the parameter space has $\Rnum>1$, and the threshold testing intensity required to control the epidemic increases --- as we move to the right within each plot in this row. However, when waiting individuals are at least partially isolated ($\theta_w>0$), we typically see the opposite effect: longer testing delays lead to \emph{less} transmission. This is because returning negative tests leads people to stop distancing; this applies both to susceptibles, and to infectious people who receive negative test results because they were sampled when susceptible (or because of imperfect test sensitivity, which we have not considered in our analysis). For small but non-zero waiting-isolation levels (e.g. $\theta_w=0.25$, second row from top), with high levels of confirmed isolation ($\theta_c \ge 0.5$), increasing testing delays initially decreases $\Delta$, then increases it as the testing delays pass a threshold value of approximately 2--5 days.
\item \fref{pan} also shows that greater testing intensity (increasing $\rho$) generally increases the effectiveness of control (moving up in each panel). This relationship can be reversed, but only for an unrealistic scenario --- very high testing intensities \emph{that begin before the epidemic starts}. It occurs when waiting people do not isolate ($\theta_w=0$), and most easily for targeted testing (\fref{pan2}(b), right three panels of top row) \bmb{can we rule out the possibility that this phenomenon could happen with random testing? In the left panel of \fref{pan2}, $\theta_w=0$/$\theta_c=0.75$ panel, there is a hint of this behaviour; it looks like the light green [second lightest] region displays this behaviour, but it's hard to see even a maximum magnification}. \ali{Yes we can rule out this for random testing, (see Eqs.~\ref{eq:dd3dr} and \ref{eq:abcd2} and the brief analysis followed)}This phenomenon occurs --- increasing testing intensity \emph{increases} \Rnum --- because more rapid testing leaves more susceptibles in the ``waiting-for-negative-results'' category at the DFE; these people must then wait for their tests to be returned (during which time they may become infected) before they can be tested again, receive a positive test, and then begin self-isolating. This effect is usually weak compared to the beneficial effects of testing.
  \bmb{I am still trying to understand why slowing testing return (decreasing $\omega$) decreases \Rnum over \textbf{most} of the parameter space considered here, which is a counterintuitive/silly result. Fast test returns help when we have high test positivity and low isolation during the waiting period (specifically, [much] lower isolation while waiting than when confirmed, but I think we're assuming that); we want to get waiting people into the ``confirmed'' compartment as quickly as possible. Can we understand/explain this result more thoroughly using that logic, and can we argue that what we are looking at is \emph{not} a realistic parameter range? Does the low baseline \Rnum we are using play into this, e.g. by reducing test positivity? Is there a simple expression for test positivity during the invasion phase?}

>>>>

%% FIXME: uncomment when this pipeline is working
\input{codes/modeldefs.tex}

@@

%% FIXME: uncomment when pipeline works
$\beta=\betaparam$; $\gamma=\gammaparam$

@@

\newpage
\begin{figure}[h!]
\centering
\begin{subfigure}[t]{.49\textwidth}
\centering
\includegraphics[width=\linewidth]{codes/R0contour_random.pdf}
\caption{}\label{p.a}
\end{subfigure}
%
\begin{subfigure}[t]{.49\textwidth}
\centering
\includegraphics[width=\linewidth]{codes/R0contour_TTI.pdf}
\caption{}\label{p.b}
\end{subfigure}
\caption{
{\bf A comparison of effectiveness control on $\Rnum$ at low level of \percap testing intensity $\rho$.}
Numerical evaluation of the effectiveness control parameter, $\Delta$ (Eq.~\ref{eq:del4}), with respect to changes in testing and isolation. Parameter values (Table~\ref{tab:params}):
$\beta=0.25$ day$^{-1}$, $1/\gamma= 1/3$ days; $\omega \in [1/12,2]$ day$^{-1}$;  $\rho \in [0,0.013]$ day$^{-1}$; $\theta\_w$ and $\theta\_c$ vary between 0 (no effect of isolation) and 1 (complete elimination of transmission); $p_S=0$, $p_I=1$ and $p_R=0.5$. Contours of $\Delta$ are plotted for (a) random testing ($w_S=w_I=w_R=1$) and (b) targeted testing ($w_S=0.3$; $w_I=w_R=1$). 
The baseline basic reproduction number $\Rnum$ and the doubling time $r$, i.e. $\beta/\gamma$ and $\beta-\gamma$ respectively, are approximately 1.5 and 0.08 in this simulation.
}
\label{pan}
\end{figure}

\begin{figure}[h!]
\centering
\begin{subfigure}[t]{.49\textwidth}
\centering
\includegraphics[width=\linewidth]{codes/R0contour_random2.pdf}
\caption{}
\end{subfigure}
%
\begin{subfigure}[t]{.49\textwidth}
\centering
\includegraphics[width=\linewidth]{codes/R0contour_TTI2.pdf}
\caption{}
\end{subfigure}
\caption{
{\bf A comparison of effectiveness control on $\Rnum$ at high level of \percap testing intensity $\rho$.}
We numerically evaluate the effectiveness control parameter, $\Delta$ (Eq.~\ref{eq:del4}), with respect to testing and isolation. We use the following parameters (see Table~\ref{tab:params} for a brief description):
$\omega \in [1/5,2]$ 1/day, $1/\gamma= 6$ days, $\rho \in [0,1/5)$ 1/day, $\theta\_w$ and $\theta\_c$ vary between 0 and 1 with 0 for no effect and 1 for full effect of isolation on the transmission probability, $\beta=0.25$ 1/day, $p_S=0$, $p_I=1$ and $p_R=0.5$. Contours of $\Delta$ are plotted for two testing strategies identified by a set of relative testing weights; (a) random testing where $w_S=w_I=w_R=1$ and (b) targeted testing where $w_S=0.3$ and $w_I=w_R=1$. The baseline basic reproduction number $\Rnum$ and the doubling time $r$, i.e. $\beta/\gamma$ and $\beta-\gamma$ respectively, are approximately 1.5 and 0.08 in this simulation.
}
\label{pan2}
\end{figure}

@@

\begin{itemize}
\item When waiting individuals do not isolate (e.g. $\theta_w =0$, top row of \fref{pan}), increasing testing delays (i.e. decreasing $\omega$) increases \Rnum, as we might expect: $\Delta$ decreases --- a larger region of the parameter space has $\Rnum>1$, and the threshold testing intensity required to control the epidemic increases --- as we move to the right within each plot in this row. However, when waiting individuals are at least partially isolated ($\theta_w>0$), we typically see the opposite effect: longer testing delays lead to \emph{less} transmission. This is because returning negative tests leads people to stop distancing; this applies both to susceptibles, and to infectious people who receive negative test results because they were sampled when susceptible (or because of imperfect test sensitivity, which we have not considered in our analysis). For small but non-zero waiting-isolation levels (e.g. $\theta_w=0.25$, second row from top), with high levels of confirmed isolation ($\theta_c \ge 0.5$), increasing testing delays initially decreases $\Delta$, then increases it as the testing delays pass a threshold value of approximately 2--5 days.
\item \fref{pan} also shows that greater testing intensity (increasing $\rho$) generally increases the effectiveness of control (moving up in each panel). This relationship can be reversed, however, at very high testing intensities. This can only occur when testing is targeted, and $\theta_w$ is relatively small (\fref{pan2}(b), right three panels of top row). It is theoretically possible for increasing testing intensity to \emph{increase} \Rnum because more rapid testing leaves more susceptibles in the ``waiting-for-negative-results'' category at the DFE; if these people become infected while waiting, they will need to wait for their negative test result before they can be tested again, receive a positive test, and then begin self-isolating. This effect is usually weak compared to the beneficial effects of testing.
